---
title: "COVID and Iris Dataset Visualization"
author: "Sushrut Gaikwad"
date: '2025-02-16'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(plotly)
library(maps)
```

# Part 1: Iris Dataset

## Q1.1: Read the `iris.csv` file

```{r}
iris_sample <- read_csv("../data/iris.csv")
spec(iris_sample)

iris_data <- read_csv("../data/iris.csv", col_types = cols(
  Sepal.Length = col_double(),
  Sepal.Width = col_double(),
  Petal.Length = col_double(),
  Petal.Width = col_double(),
  Species = col_factor()
))
```

## Q1.2: Show some values

```{r}
head(iris_data)
```

## Q1.3: Histogram with `plotly`

```{r}
plot_ly(
  data = iris_data,
  x = ~Sepal.Length,
  color = ~Species,
  type = "histogram",
  opacity = 0.6,
  xbins = list(size = 0.3)
) |> layout(
  title = "Histogram of Sepal Length for Each Species",
  xaxis = list(title = "Sepal Length"),
  yaxis = list(title = "Count"),
  barmode = "overlay"
)
```

## Q1.4: Histogram with ggplot2 and ggplotly

```{r}
gg_histogram <- ggplot(iris_data, aes(x = Sepal.Length, fill = Species)) +
  geom_histogram(alpha = 0.6, position = "identity", binwidth = 0.3) +
  theme_minimal() +
  labs(
    title = "Histogram of Sepal Length for Each Species",
    x = "Sepal Length",
    y = "Count"
  )

ggplotly(gg_histogram)
```

## Q1.5: 2x2 Facet Histogram

```{r}
iris_long <- iris_data |>
  pivot_longer(cols = c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width), 
               names_to = "Metric", 
               values_to = "Value")

metrics <- unique(iris_long$Metric)

plots <- lapply(seq_along(metrics), function(i) {
  plot_ly(
    data = filter(iris_long, Metric == metrics[i]),
    x = ~Value,
    color = ~Species,
    type = "histogram",
    opacity = 0.6,
    showlegend = (i == 1)
  )
})

subplot(plots[[1]], plots[[2]], plots[[3]], plots[[4]], 
        nrows = 2, shareX = TRUE, shareY = TRUE) |>
  layout(
    title = "Histograms of Iris Metrics by Species",
    showlegend = TRUE
  )
```

## Q1.6: Which metric separates species best?

Petal.Length clearly provides the best species separation.

## Q1.7: 2x2 Box Plot

```{r}
plots <- lapply(seq_along(metrics), function(i) {
  plot_ly(
    data = filter(iris_long, Metric == metrics[i]),
    x = ~Species,
    y = ~Value,
    color = ~Species,
    type = "box",
    showlegend = (i == 1)
  )
})

subplot(plots[[1]], plots[[2]], plots[[3]], plots[[4]], 
        nrows = 2, shareX = TRUE, shareY = TRUE) |>
  layout(
    title = "Box Plots of Iris Metrics by Species",
    showlegend = TRUE
  )
```

## Q1.8: 2D Scatter Plot

```{r}
plot_ly(
  data = iris_data,
  x = ~Petal.Length,
  y = ~Petal.Width,
  color = ~Species,
  type = "scatter",
  mode = "markers"
)
```

## Q1.9: 3D Scatter Plot

```{r}
plot_ly(
  data = iris_data,
  x = ~Petal.Length,
  y = ~Petal.Width,
  z = ~Sepal.Length,
  color = ~Species,
  type = "scatter3d",
  mode = "markers"
)
```

## Q1.10: Comment

Setosa is clearly separable, while Versicolor and Virginica overlap.

# Part 2: COVID-19 Dataset

## Q2.1: Read `us-states.csv`

```{r}
usstates_sample <- read_csv("../data/us-states.csv")
spec(usstates_sample)

usstates <- read_csv("../data/us-states.csv", col_types = cols(
  date = col_date(format = "%Y-%m-%d"),
  state = col_factor(),
  fips = col_double(),
  cases = col_double(),
  deaths = col_double()
))
```

## Q2.2: Show values

```{r}
head(usstates)
```

## Q2.3: Monthly Cases

```{r}
monthly_cases <- usstates |>
  mutate(year_month = floor_date(date, "month")) |>
  group_by(state, year_month) |>
  summarize(
    new_cases = last(cases) - first(cases),
    .groups = "drop"
  )

head(monthly_cases)
```

## Q2.4: Plot all states over time

```{r}
plot_ly(data = monthly_cases,
        x = ~year_month,
        y = ~new_cases,
        color = ~state,
        type = 'scatter',
        mode = 'lines')
```

## Q2.5: NY State

```{r}
ny_monthly_cases <- filter(monthly_cases, state == "New York")

plot_ly(data = ny_monthly_cases, 
        x = ~year_month, 
        y = ~new_cases, 
        type = 'scatter', 
        mode = 'lines+markers')
```

## Q2.6: Peak Month NY

```{r}
ny_monthly_cases |> filter(new_cases == max(new_cases))
```

## Q2.7: Choropleth map of peak month

```{r}
peak_month <- ny_monthly_cases |> filter(new_cases == max(new_cases)) |> pull(year_month)

peak_month_cases <- monthly_cases |>
  filter(year_month == peak_month) |>
  mutate(state_abb = state.abb[match(state, state.name)],
         state_abb = ifelse(state == "District of Columbia", "DC", state_abb))

plot_ly(
  data = peak_month_cases,
  locations = ~state_abb,
  locationmode = "USA-states",
  z = ~new_cases,
  type = "choropleth",
  colorscale = "Viridis"
)
```

## Q2.8: Animation

```{r}
animated_data <- monthly_cases |>
  mutate(
    year_month = as.character(year_month),
    state_abb = state.abb[match(state, state.name)],
    state_abb = ifelse(state == "District of Columbia", "DC", state_abb)
  )

plot_ly(
  data = animated_data,
  locations = ~state_abb,
  locationmode = "USA-states",
  z = ~new_cases,
  frame = ~year_month,
  type = "choropleth",
  colorscale = "Viridis"
)
```

## Q2.9: Comments

- The animated map provides temporal trends.
- Easier to compare geographically than cluttered line plots.
